# Read the data

Now that you have a script and have loaded the necessary packages, you're ready to get coding!

To read the data, you will use a function named `read_excel()`. Here we run that function and assign the results to an object names `finches`.

```{r}
# read data ----------------------------------------------------------------

# read the finches data
finches <- read_excel("finches_data.xlsx")
```


If you get an error at this point (red text in the console output) see the two warnings below.

<span class="label label-warning">Warning</span> If you didn't run the two `library()` commands, then when your run `read_excel()` you will get an error that says:

`Error in read_excel() : could not find function "read_excel"`

To solve this error, run the two `library()` commands and then retry reading the data.

<p class="text-success">To look at an object you have just created, print it to the console window by <strong>typing its name and running it</strong>:</p>

```{r print-finches}
# print the results in the console
finches
```

The first thing you notice is that this object is not a vector like the objects you were working with before. Instead, it's a **tibble**, which is an object that stores tabular data.

The first line of the output gives the data type (tibble) and dimensions of the data (number of rows x number of columns). In the tidyverse, columns are referred to as "variables".

The first 10 rows are shown, as well as the first few variables. How many variables depends on how wide your Console window is.

For each variable, it's name is given, followed by the type of data. Most of the time we worth with either numeric data (`<dbl>`) or text data (`<chr>`).

If there are more than 10 rows, or more variables than can fit on the screen, there will be a sentence at the end saying "... with xx more rows, and x more variables", followed by a list of the unshown variables.

A quick way of seeing the first few observations for all variables is to use the `glimpse()` function:

```{r}
# take a quick look at all the variables in the dataset
glimpse(finches)
```

If you want to see the entire dataset, click the name of the data object in the Environment tab, or run `View(finches)`. This will open a handy data viewer window. Once you are done looking at the data, close the window and continue.
